Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-19 paper/paper.tex
Keep listings within margins

@@

\usepackage{listings}
\lstdefinelanguage{Agda}{
alsoletter={0123\&'},
keywords={data, where, module, import, open, public,
record, field, let, in, if, then, else, case, of, with,
do, postulate, primitive, mutual, abstract, private,
forall, exists, cong, set, prop, sort, Level, Data, Type,
Renamer},
morekeywords=[2]{L, Bool, Nat, Fin},
morekeywords=[3]{String, Dyck, Trace, Exp, Atom, O, D, C, A, NotStartsWithLP, NotStartsWithRP},
morekeywords=[4]{bal, .nil, .cons, cons, nil, inl, inr, stop, 1to1, 1to2, 0to2, 0to1,
  done, add, num, parens, left, unexpected, lookAheadRP, lookAheadNot, closeGood, closeBad,
  doneGood, doneBad},
morekeywords=[5]{.states, states, .isAcc, isAcc, transitions, src, dst, .transitions, .dst,
  .src, .label, label, 0 , 1 , 2, 3, true, false},
morekeywords=[6]{NUM},
morekeywords=[7]{\&, '},
sensitive=true,
comment=[l]{--},
morecomment=[s]{\{-}{-\}},
morestring=[b]",
mathescape=true,
moredelim={[is][]{\#}{\#}},
escapeinside={(*@}{@*)},
escapebegin=\color{funarrowcolor},
escapeend={}
}
\lstset{
language=Agda,
basicstyle=\ttfamily\footnotesize,
keywordstyle=\color{keywordcolor},
keywordstyle=[2]\color{nonlintycolor},
keywordstyle=[3]\color{lintycolor},
keywordstyle=[4]\color{linconstructorcolor},
keywordstyle=[5]\color{nonlinconstructorcolor},
keywordstyle=[6]\color{alphabetcolor},
keywordstyle=[7]\color{funarrowcolor},
identifierstyle=\color{black},
commentstyle=\color{gray}\textit,
stringstyle=\color{orange},
numbers=none,
numberstyle=\tiny\color{gray},
stepnumber=1,
numbersep=5pt,
showstringspaces=false,
tabsize=4,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
rulecolor=\color{black},
texcl=true,
backgroundcolor=\color{backcolour},
frame=single
}

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
f (a , b) = inl (a , b)
\end{lstlisting}\begin{mathpar}
  \footnotesize
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \otimes \literal b) \oplus \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\(
{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
\).

@@

\begin{figure}
  \begin{minipage}[t]{.6\textwidth}
    \vspace{0pt}
  \begin{lstlisting}
data Trace : (s : Fin 3) (*@$\tocode$@*) L where
  stop : (*@$\uparrowcode$@*)(Trace 2)
  1to1 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
  1to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
  0to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
  0to1 : (*@$\uparrowcode$@*)(Trace 1 (*@$\lto$@*) Trace 0)

k : (*@$\uparrowcode$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*)  Trace 0)
k (a , b) = 0to1 (1to1 a (1to2 b stop))
\end{lstlisting}
  \end{minipage}%
  \begin{minipage}[t]{.4\textwidth}
  \vspace{8pt}
  \hspace{10pt}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (0) {$\0$};
    \node[state, below left of=0] (1) {$\1$};
    \node[state, right of=1, accepting] (2) {$\2$};

@@

$\Trace_{\N}$ is an inductive type indexed by the starting
state of the trace $\s : \N.\states$, and it may be built
through one of three constructors. We may terminate a trace at an
accepting state with the constructor $\nil$. Here we use an Agda-style
Unicode syntax for $\bigamp$, as well as using the function arrow to
mean a non-dependent version of $\bigamp$. If we had a trace beginning
at the destination state of a transition, then we may use the $\cons$
constructor to combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the
transition.  Finally, if we had a trace beginning at the destination
of an $\epsilon$-transition then we may use $\epscons$ to pull it back
along the $\epsilon$-transition and construct a trace beginning at the
source of the $\epsilon$-transition. As a shorthand, we write $Parse_{N}$
for the accepting traces out of $N.init$.
\newcommand{\D}{\texttt{D}}
\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\tocode$@*) L where
  nil : (*@$\uparrowcode$@*)(&[ s : N .states ] N .isAcc s (*@$\tocodelin$@*) Trace(*@$_{\color{black}\N}$@*) s)
  cons : (*@$\uparrowcode$@*)(&[ t : N .transitions ] ('#N .label t#' (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .dst t)
                                                 (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .src t)))
  (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrowcode$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ] (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))

data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\tocode$@*) L where
  nil : (*@$\uparrowcode$@*)(&[ s : (*@\color{black}D@*) .states ] Trace(*@$_{\color{black}\D}$@*) s (*@\color{black}D@*) .isAcc s)
  cons : (*@$\uparrowcode$@*)(&[ c : (*@$\color{nonlintycolor}\Sigma$@*) ] &[ s : (*@\color{black}D@*) .states ]
           &[ b : Bool ] ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
\end{lstlisting}
\caption{Traces of an NFA $N$ and a DFA $D$}
\label{fig:nfatrace}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
parse(*@$\color{black}_D$@*) : (*@$\uparrowcode$@*)(String (*@$\lto$@*) &[ s : (*@\color{black}D@*) .states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b)
parse(*@$\color{black}_D$@*) String .nil s = (*@$\color{linconstructorcolor}\sigma$@*) ((*@\color{black}D@*) .isAcc s) (Trace(*@$_{\color{black}\D}$@*) .nil s)
parse(*@$\color{black}_D$@*) (String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) w) s = let (*@$\color{linconstructorcolor}\sigma$@*) b t = parse w ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) in
                                    (*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c s b a t)

print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\tocode$@*) (*@$\uparrowcode$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .nil .s)) = String .nil
print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c .s) b a trace)) =
  String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) (print(*@$\color{black}_D$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) ((*@$\color{linconstructorcolor}\sigma$@*) b trace))
\end{lstlisting}
\caption{Parser/printer for DFA traces}
\label{fig:printdfa}
\end{figure}

@@

A simple example of a CFG is the Dyck grammar of balanced
parentheses, which we define in \Cref{fig:dyckinductive}.
$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
the empty string is balanced, and the $bal$ constructor builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse. We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
that $Parse_{M}$ is strongly equivalent to $\Dyck$.
\begin{figure}
\begin{lstlisting}
data Dyck : L where
  nil : (*@$\uparrowcode$@*) Dyck
  bal : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
\end{lstlisting}
\caption{The Dyck grammar as an inductive linear type}
\label{fig:dyckinductive}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrowcode$@*)('#NUM#' (*@$\lto$@*) Atom)
  parens : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  left : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '#NUM#' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] (NotStartsWithLP (*@$\lto$@*) O n false)
data D : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  lookAheadRP : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ((')' (*@$\otimes$@*) (*@$\color{lintycolor}\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  closeGood : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrowcode$@*)(')' (*@$\lto$@*) C 0 false)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] NotStartsWithRP (*@$\lto$@*) C n false)
data A : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  doneGood : (*@$\uparrowcode$@*)(A 0 true)
  doneBad : (*@$\uparrowcode$@*)(&[ n : Nat ] A (n + 1) false)
  add : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) '#NUM#') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}

>>>>

% \usepackage{showframe}

@@

\usepackage{listings}
\lstdefinelanguage{Agda}{
alsoletter={0123\&'},
keywords={data, where, module, import, open, public,
record, field, let, in, if, then, else, case, of, with,
do, postulate, primitive, mutual, abstract, private,
forall, exists, cong, set, prop, sort, Level, Data, Type,
Renamer},
morekeywords=[2]{L, Bool, Nat, Fin},
morekeywords=[3]{String, Dyck, Trace, Exp, Atom, O, D, C, A, NotStartsWithLP, NotStartsWithRP},
morekeywords=[4]{bal, .nil, .cons, cons, nil, inl, inr, stop, 1to1, 1to2, 0to2, 0to1,
  done, add, num, parens, left, unexpected, lookAheadRP, lookAheadNot, closeGood, closeBad,
  doneGood, doneBad},
morekeywords=[5]{.states, states, .isAcc, isAcc, transitions, src, dst, .transitions, .dst,
  .src, .label, label, 0 , 1 , 2, 3, true, false},
morekeywords=[6]{NUM},
morekeywords=[7]{\&, '},
sensitive=true,
comment=[l]{--},
morecomment=[s]{\{-}{-\}},
morestring=[b]",
mathescape=true,
moredelim={[is][]{\#}{\#}},
escapeinside={(*@}{@*)},
escapebegin=\color{funarrowcolor},
escapeend={}
}
\lstset{
language=Agda,
basicstyle=\ttfamily\footnotesize,
keywordstyle=\color{keywordcolor},
keywordstyle=[2]\color{nonlintycolor},
keywordstyle=[3]\color{lintycolor},
keywordstyle=[4]\color{linconstructorcolor},
keywordstyle=[5]\color{nonlinconstructorcolor},
keywordstyle=[6]\color{alphabetcolor},
keywordstyle=[7]\color{funarrowcolor},
identifierstyle=\color{black},
commentstyle=\color{gray}\textit,
stringstyle=\color{orange},
numbers=none,
numberstyle=\tiny\color{gray},
stepnumber=1,
numbersep=5pt,
showstringspaces=false,
tabsize=4,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
rulecolor=\color{black},
texcl=true,
backgroundcolor=\color{backcolour},
frame=single,
xleftmargin = 1.5mm ,
xrightmargin = 1.5mm
}

\lstnewenvironment{floatlisting}{
  \FloatBarrier
  \lstset{
    aboveskip=2mm
  }
}{}

@@

\begin{figure}
\begin{floatlisting}
f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
f (a , b) = inl (a , b)
\end{floatlisting}\begin{mathpar}
  \footnotesize
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \otimes \literal b) \oplus \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\(
{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
\).

@@

\begin{figure}
  \begin{minipage}[t]{.6\textwidth}
    \vspace{0pt}
  \begin{floatlisting}
data Trace : (s : Fin 3) (*@$\tocode$@*) L where
  stop : (*@$\uparrowcode$@*)(Trace 2)
  1to1 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
  1to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
  0to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
  0to1 : (*@$\uparrowcode$@*)(Trace 1 (*@$\lto$@*) Trace 0)

k : (*@$\uparrowcode$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*)  Trace 0)
k (a , b) = 0to1 (1to1 a (1to2 b stop))
\end{floatlisting}
  \end{minipage}%
  \begin{minipage}[t]{.4\textwidth}
  \vspace{8pt}
  \hspace{10pt}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (0) {$\0$};
    \node[state, below left of=0] (1) {$\1$};
    \node[state, right of=1, accepting] (2) {$\2$};

@@

$\Trace_{\N}$ is an inductive type indexed by the starting
state of the trace $\s : \N.\states$, and it may be built
through one of three constructors. We may terminate a trace at an
accepting state with the constructor $\nil$. Here we use an Agda-style
Unicode syntax for $\bigamp$, as well as using the function arrow to
mean a non-dependent version of $\bigamp$. If we had a trace beginning
at the destination state of a transition, then we may use the $\cons$
constructor to combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the
transition.  Finally, if we had a trace beginning at the destination
of an $\epsilon$-transition then we may use $\epscons$ to pull it back
along the $\epsilon$-transition and construct a trace beginning at the
source of the $\epsilon$-transition. As a shorthand, we write $Parse_{N}$
for the accepting traces out of $N.init$.
\newcommand{\D}{\texttt{D}}
\begin{figure}
\begin{floatlisting}
data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\tocode$@*) L where
  nil : (*@$\uparrowcode$@*)(&[ s : N .states ] N .isAcc s (*@$\tocodelin$@*) Trace(*@$_{\color{black}\N}$@*) s)
  cons : (*@$\uparrowcode$@*)(&[ t : N .transitions ] ('#N .label t#' (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .dst t)
                                                 (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .src t)))
  (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrowcode$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ] (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))
data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\tocode$@*) L where
  nil : (*@$\uparrowcode$@*)(&[ s : (*@\color{black}D@*) .states ] Trace(*@$_{\color{black}\D}$@*) s (*@\color{black}D@*) .isAcc s)
  cons : (*@$\uparrowcode$@*)(&[ c : (*@$\color{nonlintycolor}\Sigma$@*) ] &[ s : (*@\color{black}D@*) .states ]
           &[ b : Bool ] ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
\end{floatlisting}
\caption{Traces of an NFA $N$ and a DFA $D$}
\label{fig:nfatrace}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
parse(*@$\color{black}_D$@*) : (*@$\uparrowcode$@*)(String (*@$\lto$@*) &[ s : (*@\color{black}D@*) .states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b)
parse(*@$\color{black}_D$@*) String .nil s = (*@$\color{linconstructorcolor}\sigma$@*) ((*@\color{black}D@*) .isAcc s) (Trace(*@$_{\color{black}\D}$@*) .nil s)
parse(*@$\color{black}_D$@*) (String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) w) s = let (*@$\color{linconstructorcolor}\sigma$@*) b t = parse w ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) in
                                    (*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c s b a t)
print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\tocode$@*) (*@$\uparrowcode$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .nil .s)) = String .nil
print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c .s) b a trace)) =
  String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) (print(*@$\color{black}_D$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) ((*@$\color{linconstructorcolor}\sigma$@*) b trace))
\end{lstlisting}
\caption{Parser/printer for DFA traces}
\label{fig:printdfa}
\end{figure}

@@

A simple example of a CFG is the Dyck grammar of balanced
parentheses, which we define in \Cref{fig:dyckinductive}.
$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
the empty string is balanced, and the $bal$ constructor builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse. We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
that $Parse_{M}$ is strongly equivalent to $\Dyck$.
\begin{figure}
\begin{floatlisting}
data Dyck : L where
  nil : (*@$\uparrowcode$@*) Dyck
  bal : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
\end{floatlisting}
\caption{The Dyck grammar as an inductive linear type}
\label{fig:dyckinductive}
\end{figure}

@@

\begin{figure}
\begin{floatlisting}
data Exp : L where
  done : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrowcode$@*)('#NUM#' (*@$\lto$@*) Atom)
  parens : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  left : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '#NUM#' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] (NotStartsWithLP (*@$\lto$@*) O n false)
data D : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  lookAheadRP : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ((')' (*@$\otimes$@*) (*@$\color{lintycolor}\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  closeGood : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrowcode$@*)(')' (*@$\lto$@*) C 0 false)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] NotStartsWithRP (*@$\lto$@*) C n false)
data A : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
  doneGood : (*@$\uparrowcode$@*)(A 0 true)
  doneBad : (*@$\uparrowcode$@*)(&[ n : Nat ] A (n + 1) false)
  add : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) '#NUM#') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
\end{floatlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
